---
title: 'Spatial Analysis Assignment 1 '
output:
  html_document: 
    df_print: paged
---
# An analysis of Land Registry house prices in Liverpool

```{r, results='hide', quietlt = T, message=F, warning=F}
setwd("~/Documents/Geographic Data Science /Spatial analysis/Assign")
#Load required packages
library(tufte)
library(knitr)
library(tidyverse)
library(kableExtra)
library(sf) 
library(sp) 
library(tmap) 
library(RColorBrewer) 
library(viridis) 
library(rgdal)
library(ggmap)
library(GISTools)
library(gstat)
library(spdep)

```


```{r, echo=T , eval=T}
#import house price data
HouseData <- readOGR("Data/feb16.gpkg")

#Change price from factor to numeric
HouseData$price <- as.numeric(as.character(HouseData$price))


```

#### KDE 

```{r, echo=T, eval = T,  message = F}
#change co-ord system
wgs84 <- CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")

HDWgs84 <- spTransform(HouseData, wgs84)

HDWgs84@data['lon'] <- HDWgs84@coords[, 1]
HDWgs84@data['lat'] <- HDWgs84@coords[, 2]

coords <- HDWgs84@data[c("lon", "lat")]

#live bounding box
liv <- c(left = min(coords$lon)-0.05 , bottom = min(coords$lat), 
         right = max(coords$lon), top = max(coords$lat))

#basemap
basemap <- get_stamenmap(liv, zoom = 12, maptype = 'terrain', crop = T)

#basemap with kde
final <- ggmap(basemap, extent = "device", 
               maprange=FALSE) + 
  stat_density2d(data = HDWgs84@data, 
                aes(x = lon, y = lat, 
                    alpha= ..level.., 
                    fill = ..level..), 
                size = 0.01, bins = 16, 
                geom = 'polygon', 
                show.legend = F )+
  scale_fill_gradient2("Transaction\nDensity", 
                       low = "#fffff8", 
                       high = "#8da0cb")

final

```


The spatial KDE allows us to interpret the continuous spatial distribution of house transactions and perhaps more importantly relate this to the underlying geography. For this period there is a high probability estimate for house transactions in the south and south-west of the city, with particularly high concentrations represented by darker colours, and very few in the city centre. 




### Spatial interpolation

Whilst KDE's can give us a sense of the distribution of house transactions they fail to give any more interesting information on the nature of these transactions and how certain characteristics i.e. The price of houses being sold, may vary across space. For this purpose, we can make use of spatial interpolation in order to create a continuous surface representing house prices. Specifically using the inverse distance weighting approach in order to provide house price estimates for a grid of points across the geography we are interested in, this is done using the available observations and weighting them by distance to provide estimates for all points in out grid, in this case we used the closest 75 houses from each point.


```{r, message = F, results=F }
#Grid of points
liv.grid <- spsample(HDWgs84, type='regular', n=100000)
#create idw estimates for price based on 75 nearestn
idw.hp <- idw(price ~ 1, locations=HDWgs84, newdata=liv.grid, nmax = 75)

```

```{r, message = F}
#get estimates and co-ords
xyz <- data.frame(x=coordinates(idw.hp)[, 1], 
                  y=coordinates(idw.hp)[, 2], 
                  z=idw.hp$var1.pred)
                  

#get log house prices
xyz['lz'] <- log(xyz$z)

#plot surface as raster
base <- ggplot(data=xyz, aes(x=x, y=y))
surface <- base + geom_raster(aes(fill=lz),
                       show.legend = F) +
  scale_fill_gradient(low = "white", high = "blue")
surface


```

Given the skewed nature of house prices here the estimate for $log(House Price)$ is plotted as a surface. Allowing use to see the distribution of houses prices across space.

```{r, message = F, warning = F,results=F}
liv <- c(left = min(coords$lon) , bottom = min(coords$lat), 
         right = max(coords$lon), top = max(coords$lat))
basemap <- get_stamenmap(liv, zoom = 12, maptype = 'terrain', crop = T)

#plot basemap and surface overlaid 
surface2 <- ggmap(basemap, base_layer = ggplot(xyz, aes(x=x,y=y, z=lz))) + 
    geom_raster(aes(fill=lz), alpha = 0.8) + coord_cartesian()+
    scale_fill_gradient(low = "white", high = "blue")+
    
    labs( x = 'Longitude', y = 'Latitude', fill= "log(House price)")

surface2
```

As we did previously It would be useful to understand the underlying geography in more detail. This happens to be practically a little difficult to get right in this situation. However, some context can still be gained. 


### Non-spatial regression

We have seen how house prices can vary over space, so now we seek to find some explanation for the price differences. In particular we wish to explain the variation in the house prices as a function of whether it is new and the Index of Multiple deprivation of the area. So we have $$log(P_i) = \alpha + \beta_1NEW_i + \beta_2IMD_i + \epsilon_i$$

log house prices are used here so that coefficients can be interpreted as a percentual change induced by a unit increase.



```{r}
m2 <- lm('log(price) ~ new + imd_score', HouseData)
summary(m2)

```

Fitting this linear model we see that a brand new house is estimated to be a staggering 43% more expensive compared to an old house. A unit increase in deprivation reduces a house value by around 2%


### Spatial FE

The previous regression model did not take into account any geographical context. Now we wish to introduce a spatial element. This can be useful when the process we wish to explore exhibits spatial heterogeneity, that is, our outcome of interest (house price) has an underlying relationship with geography meaning prices behave differently in different locations. This is beneficial as a model with only two explanatory variables is likely to be missing other important predictive factors. This can be accounted for using a spatial fixed effects model, here we allow the regression constant $\alpha$ to vary by postcode, allowing us to control for unobserved effects that vary by location.


```{r}
HouseData$pc <- as.character(lapply(strsplit(as.character(HouseData$pcds), split=" "), "[", 1))

m3 <- lm('log(price) ~ pc + new + imd_score - 1', HouseData)
summary(m3)


```

Fitting a FE model variations are only derived from within each postcode, here we see the newness of a house and unit changes in deprivation score have a lesser effect on house prices when compared only to houses from the same postcode. FE models don't however allow the explanatory variables to vary by postcode. 


### Spatial dependence 

We have considered spatial effects in the FE model however we have not considered spatial configurations. For example, the price of a house may not only be affected by the deprivation level of the area it's in but also by the deprivation of surrounding areas. Spatial dependence can be introduced using spatial weights matrices representing the degree of spatial connectivity between observations, in this case we are dealing with point observations so will use knn weighting with $k = 75$ this will allow for enough variation and ability to pick up houses from different regions while not exhausting our quite small dataset. For each observation this will assign a value of 1 to the $k$ nearest neighbours and 0 to all other observations. 

We can now fit a regression model and in addition to including the level of deprivation in the area of the house, we will include its spatial lag which is the average deprivation of the $k$ neighbours so not only the level of deprivation of the area where a house is located but also that of the surrounding locations that will try to explain the final price at which a house is sold.



```{r}
xy.jit <- jitter(HouseData@coords)

# Create knn list of each house
hnn <- knearneigh(xy.jit, k=10)

# Create nb object
hnb <- knn2nb(hnn)


# Create spatial weights matrix (note it row-standardizes by default)
hknn <- nb2listw(hnb)

#Add spatial lag IMD scores 
HouseData@data$w_imd_score <- lag.listw(hknn, HouseData@data$imd_score)
```

```{r}
#Spatial dependance regression model
m6 <- lm('log(price) ~ new + imd_score + w_imd_score', HouseData)
summary(m6)
```

Fitting the model taking into account the spatial lag of IMD i.e. deprivation of surrounding areas, we see that it has a negative effect on house price as expected however not as significant as the deprivation of the house location.

### Comparison 

The dataset examined here is much smaller than the in-lab dataset so may be less representative of the average. Differences can certainly be seen between the results, firstly the distribution of house transactions is markedly different with here far less transactions observed in the city centre and far more in south Liverpool. Spatial interpolation using in lab data did a small concentration of high house prices in the south east of the city, something that can also be observed with this data set. In terms of spatial econometrics both in lab and here it was found that a house being new had a greater effect on value that the deprivation of the location, both with and without fixed effects of postcode. A notable difference in lab was the finding that surrounding area deprivation had a greater effect that location deprivation, the opposite to what was found here, it should be noted however that IMD scores are generated on a LSOA level thus the IMD score for a house location is actually also representative of a small surrounding area, with often several houses having the same IMD score.


